---
title: "Drug Toxicity Prediction with Deep Learning"
author: "Grace Bianchi"
date: "June 2026"
output:
    html_document:
      toc: true
      toc_float: true
      code_folding: show
---

# Introduction

This project applies deep learning to computational toxicology, using the Tox21 dataset to predict whether chemical compounds are toxic across 12 different biological assays. Each compound is represented by a molecular fingerprint derived from its SMILES string, and the goal is to build a multi-task neural network capable of predicting toxicity across all 12 endpoints simultaneously. 

## SMILES Background

SMILES (Simplified Molecular Input Line Entry System) is a notation for representing chemical structures as text strings that computers can parse and process. Atoms are denoted by their elemental symbol, with case indicating aromaticity (uppercase for non-aromatic, lowercase for aromatic), and bonds are represented by symbols such as `=` for double bonds and `#` for triple bonds, with single bonds left implicit. Molecules are hydrogen-suppressed by default, meaning hydrogens are inferred automatically rather than written out explicitly. Branching structures are enclosed in parentheses, ring closures are marked with paired numbers, and charged atoms are specified using brackets (e.g. `[O-]`). Together these rules allow complex molecular structures to be compactly encoded in a single line of text, which is what makes SMILES strings usable as model input after being converted to molecular fingerprints.

# Exploratory Data Analysis

## Loading and Tidying Data

We begin by loading the packages needed for data manipulation, modeling, and visualization. 

```{r, message = F, warning = F}
# Loading Libraries
library(tidyverse)
library(tidymodels)
library(naniar)
library(torch)
library(pROC)
library(factoextra)
library(FactoMineR)
```

SMILES strings were converted to 2,048-bit Morgan molecular fingerprints using RDKit in Python and saved as a CSV, which is loaded here alongside the original compound data and toxicity labels.

```{r, warning = F}
# Converted smiles to fingerprints in python, loading that csv here
tox21_w_fps <- read.csv("data/tox21_with_fingerprints.csv")
```

## Cleaning and Preprocessing

The dataset contains 7,823 compounds across 2,062 columns: 12 toxicity endpoint labels, the original SMILES string, and 2,048 binary fingerprint features. The original Tox21 training set contains 8,014 compounds. Following SMILES parsing in Python, 191 compounds were dropped due to invalid or unparseable SMILES strings, resulting in a final dataset of 7,823 compounds. Missing values are present across all 12 endpoints (7–26%), reflecting compounds that were not tested in every assay rather than confirmed non-toxic outcomes. Missing labels will be masked during model training rather than imputed to avoid introducing false negatives into the training signal.

```{r, warning = F}
dim(tox21_w_fps)

endpoints <- tox21_w_fps %>%
  select("SR.MMP", "NR.Aromatase", "SR.ARE", "NR.ER",
         "NR.PPAR.gamma", "SR.HSE", "NR.AhR", "NR.AR.LBD", 
         "SR.p53", "NR.ER.LBD", "SR.ATAD5", "NR.AR")


summary(endpoints)

miss_var_summary(endpoints)

gg_miss_var(endpoints)
```

## Visual EDA

### Toxicity Rate by Endpoint

```{r, warning = F}
endpoint_cols <- c("NR.AR", "NR.AR.LBD", "NR.AhR", "NR.Aromatase", 
                    "NR.ER", "NR.ER.LBD", "NR.PPAR.gamma", 
                    "SR.ARE", "SR.ATAD5", "SR.HSE", "SR.MMP", "SR.p53")

long_data <- tox21_w_fps %>%
  select(all_of(endpoint_cols)) %>%
  pivot_longer(everything(), names_to = "endpoint", values_to = "label") %>%
  filter(!is.na(label)) %>%
  mutate(label = factor(label, levels = c(0, 1), labels = c("Non-toxic", "Toxic")))

toxic_pct <- long_data %>%
  group_by(endpoint) %>%
  summarise(pct_toxic = mean(label == "Toxic") * 100)

ggplot(toxic_pct, aes(x = reorder(endpoint, pct_toxic), y = pct_toxic)) +
  geom_col(fill = "lightblue") +
  coord_flip() +
  labs(title = "Toxicity Rate by Endpoint",
       x = NULL, y = "% Toxic") +
  theme_minimal()
```

Toxicity rates vary substantially across endpoints, with stress response assays SR.ARE and SR.MMP showing the highest active rates (~16%), followed by nuclear receptor endpoints NR.ER and NR.AhR (~12–13%). The remaining eight endpoints fall below 7%, with NR.PPAR.gamma showing the lowest toxicity rate overall. This imbalance is consistent across all endpoints and will be addressed through class weighting during model training

### Principal Component Analysis of Molecular Fingerprints

```{r, warning = F}
fp_matrix <- tox21_w_fps %>%
  select(starts_with("fp"))

pca_result_fp <- prcomp(fp_matrix, scale = F)

fviz_eig(pca_result_fp, addlabels = TRUE)
```

The scree plot shows no clear elbow — variance explained drops gradually from 4.1% at PC1 to 1.1% by PC10, with no dominant components emerging. This flat, diffuse structure is expected for high-dimensional binary fingerprint data, where chemical information is spread thinly across 2,048 sparse bits rather than concentrated in a few correlated features. The absence of dominant principal components motivates the use of a neural network, which can learn nonlinear combinations of fingerprint features that PCA cannot capture through linear projection alone.

### PCA of Molecular Fingerprints Colored by SR.p53 Toxicity Label

```{r, warning = F}
fp.pca = PCA(fp_matrix,scale.unit = F , ncp = 6 , graph=F)

fviz_pca_ind(fp.pca, 
             geom = "point", 
             habillage = tox21_w_fps$SR.p53,
             alpha.ind = 0.5)
```

Toxic (1) and non-toxic (0) compounds show no meaningful separation along the first two principal components, with SR.p53-active compounds scattered throughout the same space as inactive ones. This confirms that toxicity is not linearly separable in fingerprint space, at least along the dimensions of greatest variance — further supporting the need for a nonlinear model capable of learning more complex decision boundaries.

### Correlation Matrix of Toxicity Endpoints

```{r, warning = F}
endpoints %>% 
  cor(use = "pairwise.complete.obs") %>%
  corrplot::corrplot(method = "color")
```

Toxicity labels show modest positive correlations across endpoints, with no strong negative correlations present. The most notable pattern is the high correlation between NR.AR and NR.AR.LBD, which is expected since both assays measure androgen receptor activity — one at the full receptor level and one specifically at the ligand binding domain. NR.ER and NR.ER.LBD show a similar relationship for the same reason. These within-pathway correlations suggest shared biological mechanisms across certain endpoint pairs, which the multi-task architecture is well-suited to exploit through shared hidden layers.

### Compound Toxicity Burden Across Endpoints

```{r, warning = F}
tox21_w_fps %>%
  mutate(total_toxic = rowSums(select(., all_of(endpoint_cols)) == 1, na.rm = TRUE)) %>%
  ggplot(aes(x = total_toxic)) +
  geom_bar(fill = "darkred", color = "black") +
  labs(title = "Number of Endpoints Triggered per Compound",
       x = "Total Toxic",
       y = "Count") +
  theme_minimal()
```

The majority of compounds (~5,000) show no active toxicity across any endpoint, while counts drop sharply as the number of triggered endpoints increases. A small number of compounds trigger 5 or more endpoints, representing "promiscuous" toxic compounds that interfere with multiple biological pathways simultaneously. This heavy right-skew reinforces the class imbalance challenge and highlights that truly multi-endpoint toxic compounds are rare but potentially the most important to correctly identify.

# Setting Up Model

## Train/Test Split

We split the dataset into training and testing sets using an 80/20 proportion, stratified by SR.ARE. Since the dataset has 12 toxicity endpoints, perfect multi-label stratification is not straightforward — instead we stratify on SR.ARE as a proxy, as it has the highest toxicity rate (~16%) and therefore the most signal to preserve proportionally across splits. Stratification ensures the class balance observed in the full dataset is maintained in both the training and testing sets. This gives us 6,258 observations for training and 1,565 for testing.

```{r, warning = F}
# Split data into training and testing
set.seed(123)

tox_split <- initial_split(tox21_w_fps, prop = 0.80, 
                               strata = SR.ARE)

tox_train <- training(tox_split)

tox_test <- testing(tox_split)

dim(tox_train)
dim(tox_test)
```

## Computing Class Weight

Class weights are computed per endpoint as the ratio of non-toxic to toxic compounds in the training set. These weights will be passed to the loss function during model training to penalize misclassification of the minority toxic class more heavily, preventing the model from defaulting to predicting non-toxic for all compounds.

```{r}
class_weights <- tox_train %>%
  select(all_of(endpoint_cols)) %>%
  summarise(across(everything(), ~ sum(. == 0, na.rm = TRUE) / sum(. == 1, na.rm = TRUE)))

class_weights
```

## Feature Scaling

Molecular fingerprints are binary vectors (0/1) where each bit indicates presence or absence of a molecular substructure. Since all features are already on the same scale by construction, normalization is not required and would risk distorting the sparse structure of the fingerprint representation.

## Converting to Tensors

Before building the model, the fingerprint features and toxicity labels are converted from data frames into torch tensors, the data structure the model operates on. Features are stored as a compounds x 2048 matrix and labels as a compounds x 12 matrix, with missing labels preserved as `NaN` so they can be masked during loss computation.

```{r}
# Extracting fingerprint and label matrices
fp_train <- tox_train %>%
  select(starts_with("fp")) %>%
  as.matrix()

fp_test <- tox_test %>%
  select(starts_with("fp")) %>%
  as.matrix()

lab_train <- tox_train %>%
  select(all_of(endpoint_cols)) %>%
  as.matrix()

lab_test <- tox_test %>%
  select(all_of(endpoint_cols)) %>%
  as.matrix()
```

```{r}
# Convert to Torch Tensors
fp_train_t <- torch_tensor(fp_train, dtype = torch_float())

fp_test_t <- torch_tensor(fp_test, dtype = torch_float())

lab_train_t <- torch_tensor(lab_train, dtype = torch_float())

lab_test_t <- torch_tensor(lab_test, dtype = torch_float())

fp_train_t$shape
lab_train_t$shape
```

## Building Dataset and Dataloader

```{r}
tox_dataset <- dataset(
  name = "tox_dataset",
  
  initialize = function(features, labels) {
    self$features <- features
    self$labels <- labels
  },
  
  .getitem = function(i) {
    list(
      x = self$features[i,],
      y = self$labels[i,]
    )
  },
  
  .length = function() {
    self$features$shape[1]
  }
)
```

```{r}
# instantiate for training and testing
train_ds <- tox_dataset(fp_train_t, lab_train_t)
test_ds <- tox_dataset(fp_test_t, lab_test_t)

# wrap in dataloaders
train_dl <- dataloader(train_ds, batch_size = 64, shuffle = TRUE)
test_dl <- dataloader(test_ds, batch_size = 64, shuffle = FALSE)
```

```{r}
batch <- train_dl$.iter()$.next()
batch$x$shape
batch$y$shape
```

